Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-08-12 Manuscript/Supplementary_Information.tex
fix figure paths

The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) of \citet{Hartmann2012a}. The raw data takes the form of polygon vectors, where each polygon is assigned one of 16 lithologic categories. We first group these 16 categories into 6 broader categories (metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment; Table \ref{tab:GLiM_to_GEOCLIM}). We then rasterize the polygon vectors to 0.1\degrees $\times$ 0.1\degrees resolution, where each pixel is assigned the lithologic category of the polygon that covers the greatest area in that pixel (i.e. the `mode lithology'; Fig. \ref{fig:world_lithology}). To improve the computing time of GEOCLIM, we decrease the resolution of the raster to 0.5\degrees $\times$ 0.5\degrees. To do so, a 3-dimensional 720 $\times$ 360 $\times$ 7 matrix is created, in which the fraction of each 0.5\degrees $\times$ 0.5\degrees pixel covered by each of the 6 lithologic categories (or water/ice) is captured by the extra dimension. In this way, we calculate an area-weighted mean Ca+Mg concentration of the surface in each 0.5\degrees $\times$ 0.5\degrees pixel.

The Ca+Mg concentrations of felsic (1,521~mol/m$^{3}$), intermediate (4,759~mol/m$^{3}$), and mafic (10,317~mol/m$^{3}$) lithologies are assigned based on the mean of data compiled from EarthChem (\url{www.earthchem.org/portal}; calculations made in the code within the repository). For metamorphic and siliciclastic lithologies, we explore a range of feasible Ca+Mg concentrations during calibration of the silicate weathering component of GEOCLIM (\textit{GEOCLIM Calibration}).

@@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/lithology_areas.pdf}
    \caption{Total area of the lithologic categories within the SEAI for each tested scenario. Note that the grey bars in the middle panel indicate the fractional area of each lithologic category for the entire Earth.}
    \label{fig:lithology_areas}
\end{figure}

@@

Proxy-based \pCOtwo estimates for the past 20~m.y. are shown in Figure \ref{fig:pCO2_proxies} \citep{Bereiter2015a, Foster2017a, Ji2018a, Cui2020a}. Even when stringent quality criteria and the latest understanding of each of the \pCOtwo proxies have been applied to available \pCOtwo records, both significant uncertainty in the estimated \pCOtwo for any given data point as well as disagreement between techniques remain. Statistical methods that are often utilized in an attempt to extract trends from this data (e.g. locally-weighted scatter plot smoothing, LOWESS) evaluate the mean (with some weighting) of the highly scattered data in a given time interval. However, such methods are only appropriate for estimating the ``true'' value at any given time interval in datasets in which individual data points are being drawn from a single distribution about a mean value. The \pCOtwo proxy compilation shown in Figure \ref{fig:pCO2_proxies} consists of several distinct techniques and samples, each with their own set of assumptions and associated probability distributions that influence their reported values. Therefore, simply taking the mean of these values neglects the fact that some of these techniques/samples are likely more reliable than the others. Constraining how robust each \pCOtwo proxy technique is remains an important challenge, and as such the ``true'' \pCOtwo could plausibly lie anywhere within the full range of proxy-based \pCOtwo estimates for any given time interval. Nevertheless, our modeled \pCOtwo values for 15~Ma (Fig. \ref{fig:scenario_pCO2_all}) resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene. Given this scatter in the proxy-based \pCOtwo estimates, we instead infer the Neogene cooling trend from the Miocene benthic foram oxygen isotope record shown in Figure 1 in the main text.

>>>>

The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) of \citet{Hartmann2012a}. The raw data takes the form of polygon vectors, where each polygon is assigned one of 16 lithologic categories. We first group these 16 categories into 6 broader categories (metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment; Table \ref{tab:GLiM_to_GEOCLIM}). Note that the siliciclastic sediment lithologic category also includes sedimentary sequences in which any carbonate is identified but is not the dominant lithology \cite{Hartmann2012a}. We then rasterize the polygon vectors to 0.1\degrees $\times$ 0.1\degrees resolution, where each pixel is assigned the lithologic category of the polygon that covers the greatest area in that pixel (i.e. the `mode lithology'; Fig. \ref{fig:world_lithology}). To improve the computing time of GEOCLIM, we decrease the resolution of the raster to 0.5\degrees $\times$ 0.5\degrees. To do so, a 3-dimensional 720 $\times$ 360 $\times$ 7 matrix is created, in which the fraction of each 0.5\degrees $\times$ 0.5\degrees pixel covered by each of the 6 lithologic categories (or water/ice) is captured by the extra dimension. In this way, we calculate an area-weighted mean Ca+Mg concentration of the surface in each 0.5\degrees $\times$ 0.5\degrees pixel.

The Ca+Mg concentrations of felsic (1,521~mol/m$^{3}$), intermediate (4,759~mol/m$^{3}$), and mafic (10,317~mol/m$^{3}$) lithologies are assigned based on the mean of data compiled from EarthChem (\url{www.earthchem.org/portal}; calculations made in the code within the repository). For metamorphic and siliciclastic lithologies, we explore a range of feasible Ca+Mg concentrations during calibration of the silicate weathering component of GEOCLIM (\textit{GEOCLIM Calibration}). This approach makes the simplifying assumption that all pixels of a given lithologic category share the same Ca+Mg concentration.

@@

A related issue is that our analysis focuses on the weathering of Ca and Mg silicates. Sources of alkalinity to the ocean include the four major cations of Ca, Mg, Na and K. A common approach for the long-term carbon cycle follows that of \citet{Berner1984a} where an emphasis is placed on Ca and Mg with the weathering of Na and K silicates considered to not be a long-term sink of \COtwo. The reasoning behind this assumption is that while weathering of these silicates can lead to transient uptake of \COtwo, this uptake is balanced by release. The ultimate mechanism through which silicate weathering sequesters carbon is the incorporation of Ca+Mg carbonates into the lithosphere. While alkalinity generated by the weathering of silicate Na and K consumes \COtwo, that alkalinity is removed through processes such as reverse weathering reactions and exchange for Ca that ultimately precipitates as carbonate \cite{Sun2016a}. An alternative approach for estimating carbon sequestration associated with silicate weathering that sought to incorporate all four cations was taken by \citet{France-Lanord1997a} who used a formulation of $\Delta CO_2 = \Delta Mg + \Delta Ca + 0.15 \Delta Na + 0.1 \Delta K$. This relationship, also applied by \citet{Schopka2011a}, assumes that all Ca$^2+$ and Mg$^2+$ eventually forms carbonates and applies an estimate that 20\% of K$^+$ and 30\% of Na$^+$ exchanges for Ca and Mg and ultimately produces carbonate (and considers charge balance). Overall, uncertainty in the sinks for K and Na and the processes that control them make them difficult to parameterize and model. Given that we are modeling steady-state solutions and the ultimate sequestration of carbon through carbonate formation, we make the simplification of restricting the modeling of carbon sequestration to that associated with Ca and Mg delivery. Different parameterizations such as that of \citet{France-Lanord1997a} could modulate our estimates, but the overall results would be similar.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/lithology_areas.pdf}
    \caption{Total area of the lithologic categories within the SEAI for each tested scenario. Note that the grey bars in the middle panel indicate the fractional area of each lithologic category for the entire Earth.}
    \label{fig:lithology_areas}
\end{figure}

@@

Proxy-based \pCOtwo estimates for the past 20~m.y. are shown in Figure \ref{fig:pCO2_proxies} \citep{Bereiter2015a, Foster2017a, Ji2018a, Cui2020a}. Even when stringent quality criteria and the latest understanding of each of the \pCOtwo proxies have been applied to available \pCOtwo records, both significant uncertainty in the estimated \pCOtwo for any given data point as well as disagreement between techniques remain. Statistical methods that are often utilized in an attempt to extract trends from this data (e.g. locally-weighted scatter plot smoothing, LOWESS) evaluate the mean (with some weighting) of the highly scattered data in a given time interval. However, such methods are only appropriate for estimating the ``true'' value at any given time interval in datasets in which individual data points are being drawn from a single distribution about a mean value. The \pCOtwo proxy compilation shown in Figure \ref{fig:pCO2_proxies} consists of several distinct techniques and samples, each with their own set of assumptions and associated probability distributions that influence their reported values. Therefore, running a regression through these values neglects the fact that some of these techniques/samples are likely more reliable than the others. Constraining how robust each \pCOtwo proxy technique is remains an important challenge, and as such the ``true'' \pCOtwo could plausibly lie anywhere within the full range of proxy-based \pCOtwo estimates for any given time interval. Nevertheless, our modeled \pCOtwo values for 15~Ma (Fig. \ref{fig:scenario_pCO2_all}) resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene. Given this scatter in the proxy-based \pCOtwo estimates, we instead infer the Neogene cooling trend from the Miocene benthic foram oxygen isotope record shown in Figure 1 in the main text.
